---
title: "Bivariate Data"
author:
  - email: "mpetevap@uni-sofia.bg"
    name: "Monika Peteva Petkova"
date: "`r format(Sys.time(), '%Y')`"
output:
  html_notebook:
  toc: yes
word_document: 
  toc: true
html_document:
  toc: true
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The relationship bertween 2 variables:

# Bivariate categorical data

#### Example 1:
Students who smoke study less amount:
* 1  - less than 5 hours studing, 
* 2  - 5 - 10 hours studying,
* 3  - more than 10 hours studying)
```{r}
smokes = c("Yes","No","No","Yes","No","Yes","Yes","Yes","No","Yes")
amount = c(1,2,2,3,3,1,2,1,3,2)
table(smokes,amount)
```

```{r}
prop.table(table(smokes,amount))
prop.table(table(smokes,amount),1)
prop.table(table(smokes,amount),2)
options(digits = 3)
```

#### Example 2:
```{r}
hair = c("blond", "blond", "black", "blond", "brown", "brown", 
         "brown", "brown", "black", "brown", "black", "brown",
         "black", "black", "black", "brown", "brown", "brown",
         "brown", "brown", "black", "brown", "black", "brown",
         "blond", "blond", "black", "blond", "brown", "brown", 
         "brown", "brown", "black", "brown", "black", "brown",
         "brown", "brown", "black", "brown", "black", "brown",
         "blond", "blond", "black", "blond", "brown", "brown")
eyes = c("blue", "green", "brown",  "blue", "green", "brown",
         "brown", "black", "black", "green", "brown", "brown",
         "green", "black", "black", "brown", "brown", "black",
         "green", "black", "black", "brown", "brown", "black",
         "brown",  "blue", "green", "brown", "brown", "black",
         "black", "green", "brown",  "blue", "green", "brown",
         "brown", "black", "brown",  "blue", "green", "brown",
         "blue", "green", "brown",  "blue", "green", "brown")
table(hair, eyes)
prop.table(table(hair, eyes))
prop.table(table(hair, eyes),1)
prop.table(table(hair, eyes),2)
```

Барплот
```{r}
barplot(table(smokes,amount))
barplot(table(amount,smokes))
barplot(table(smokes,amount),
        names.arg = c("<5 hours", "5-10 hours", ">10 hours"),
        beside = TRUE,
        col = c("Green", "Grey"),
        xlab = "study hours")
barplot(table(smokes,amount),
        names.arg = c("<5 hours", "5-10 hours", ">10 hours"),
        legend.text = TRUE,
        beside = TRUE,
        col = c("Green", "Grey"),
        xlab = "study hours")
barplot(table(amount,smokes),
        legend.text = c("<5 hours studying","5-10 hours studying",">10 hours studying"),
        beside = TRUE,
        col = c("Yellow", "Orange", "Red"))
```

You can also see:

* apply, tapply, mapply, sapply

You can also see: 

* lattice, ggplot2

# Categorical vs Numerical data
#### Example 1:
Drug test - experimental group and control group
```{r}
experimentalGroup = c(5, 5, 5, 13, 7, 11, 11, 9, 8, 9)
controlGroup = c(11, 8, 4, 5, 9, 5, 10, 5, 4, 10)
boxplot(experimentalGroup, controlGroup)
```
#### The same example:
```{r}
amount = c(5, 5, 5, 13, 7, 11, 11, 9, 8, 9, 11, 8, 4, 5, 9, 5, 10, 5, 4, 10)
category = c(1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2)
boxplot(amount ~ category)
```
breaking up the values in amount, by the categories in category\
amount by category

# Numerical vs numerical data
### Compare the distributions
#### Example
Home data
* boxplot
```{r}
library("UsingR")
head(home)
names(home)
home$old
home$new
boxplot(home$new)
boxplot(home$old)
boxplot(home$new, home$old)
mean(home$new)
mean(home$old)
sd(home$new)
sd(home$old)
scale(home$old)
scale(home$new)
boxplot(scale(home$new))
boxplot(scale(home$old))
par(mfrow = c(1, 2))
boxplot(scale(home$new), main = "new")
boxplot(scale(home$old), main = "old")
```
* Dot plots (Точкова диаграма)
```{r}
stripchart(home$old)
stripchart(home$new)
par(mfrow = c(2, 1))
stripchart(scale(home$old), main = "old")
stripchart(scale(home$new), main = "new")
```
* Violin Plot instead of boxplots
```{r}
violinplot(scale(home$old), scale(home$new))
simple.violinplot(scale(home$old), scale(home$new))
```

### Compare relationships
Investigate one numerical variable against another

* Scatter plot
```{r}
home
plot(home$old, home$new)
min(home$old)
min(home$new)
```
Entire data set
```{r}
homedata
plot(homedata$y1970, homedata$y2000)
```
Strong linear trend

* **Pearson correlation coefficients** [-1 < Corr < 1]

- Scaled Covariance between X and Y
- How one variable varies as the other does

* * 1 - strong linear relationship
* * 0 - weak one (no **linear** realtionship)

The Pearson correlation coefficient indicates the strength of a linear relationship between two variables, but its value generally does not completely characterize their relationship.\
See [picture1](https://en.wikipedia.org/wiki/Correlation_and_dependence#/media/File:Correlation_examples2.svg) and [picture2](https://en.wikipedia.org/wiki/Correlation_and_dependence#/media/File:Anscombe%27s_quartet_3.svg) from [here:](https://en.wikipedia.org/wiki/Correlation_and_dependence)
```{r}
cor(home$old, home$new)
cor(home$old, home$new)^2
```
Funny online games to predict correlations:
[game1](http://guessthecorrelation.com/)
[game2](http://istics.net/Correlations/)

```{r}
homedata
plot(homedata$y1970, homedata$y2000)
plot(homedata$y2000 ~ homedata$y1970)
cor(homedata$y1970, homedata$y2000)
plot(homedata)
```

# Linear regression
```{r}
homedata
plot(homedata$y1970, homedata$y2000)
abline(lm(homedata$y2000 ~ homedata$y1970))
simple.lm(homedata$y1970, homedata$y2000)
lm.res = simple.lm(homedata$y1970, homedata$y2000)
coef(lm.res)
coef(lm.res)[1]
coef(lm.res)[2]
simple.lm(homedata$y1970, homedata$y2000, show.residuals = TRUE)
```

```{r}
summary(lm(homedata$y2000 ~ homedata$y1970))
```

```{r}
head(emissions)
plot(emissions)
plot(emissions$CO2, emissions$GDP)
```

```{r}
florida
plot(florida)
names(florida)
cor(florida$BUSH, florida$BUCHANAN)
plot(florida$BUSH, florida$BUCHANAN)
simple.lm(florida$BUSH, florida$BUCHANAN)
simple.lm(florida$BUSH, florida$BUCHANAN, show.residuals = TRUE)
```
Strong linear relationship, except for two outliers\
Copy the code on the console and click on the two outliers.\
Esc to finish and print the identified cordinates
```{r}
simple.lm(florida$BUSH, florida$BUCHANAN)
identify(florida$BUSH, florida$BUCHANAN)
```
```{r}
florida[50,]
florida.new = florida[-50, ]
lm.res = simple.lm(florida.new$BUSH, florida.new$BUCHANAN)
coef(lm.res)[1]
coef(lm.res)[2]
x = 152846
y = coef(lm.res)[1] + coef(lm.res)[2] * x; y
```
We expect Buchanan to have received 598 votes, not 3407 as actually received.

```{r}
simple.lm(florida$BUSH, florida$BUCHANAN)
abline(65.6, 0.00348)
```
