---
title: "Introduction to R"
author:
 - email: "mpetevap@uni-sofia.bg"
   name: "Monika Peteva Petkova"
date: "`r format(Sys.time(), '%Y')`"
output:
  html_notebook:
    toc: yes
  word_document: 
    toc: true
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Basic Syntax
## Basic mat
```{r}
2 + 3
2 + 3 + 4
2 - 3
2 * 3
2 / 3
(2 + 3) * 4
2^3
```
## Comments
```{r}
# Comment
```
## Case sensitive
## Naming convention is with ".", not with "_"

## Not typified
```{r}
a = 5
a = 5.4
a = "string"
```
## Variable assignment operators
```{r}
a = 5
a <- 5

a = b = 5
a <- b <- 5

5 -> a # This is not usually used
```
## Print the value of an object
```{r}
a = 5
a
print(a) # This is not usually used
```
## Removing variable
```{r}
rm(a)
```
## Getting help
```{r}
?print
help(print)
```
## Look for function using part of its name
```{r}
apropos("mea")
```


# Data Types
* Logical
* Integer
* Numeric
* <span style="color:lightgrey">Complex</span>
* Character
* <span style="color:lightgrey">Dates</span>


**class()** &nbsp; - &nbsp; Take the type of the object\
**is.\<type\>()** &nbsp; - &nbsp; Check the \<type\> of the object\
**as.\<type\>()** &nbsp; - &nbsp; Change the \<type\> of the object\

## Logical
Represent data that can be either **TRUE** or **FALSE**.\
**True** is the same as **1**. **False** is the same as **0**.\
```{r}
x = 1
y = 2
z = x > y

# Prints the class type
class(z)

# Check if the class type is logical
is.logical(z)
# Check if the class type is logical
is.logical(2.5)
# Check if the class type is logical
is.logical(y)

# Converts y to type logical
k = as.logical(1)
k
class(k)
# Converts y to type logical
t = as.logical("string")
t
class(t)
# Converts y to type logical
f = as.logical(0)
f
class(f)
```

Some more examples:
```{r}
u = TRUE
v = FALSE

z = T
y = F

u & v
u | v
!u

2 == 3
2 != 3
2 < 3
2 <= 3

5 * TRUE

"data" == "statistics"
"data" < "statistics"

help("&") # Getting help
```

## Integer
If you want one variable to be integer, you must set it explicitly.
```{r}
x = 3 # By default it is numeric
x
class(x) # numeric
is.integer(x)

# Defines integer
y = as.integer(3)
y
class(y)
is.integer(y)

# Defines integer
z = 3L
z
class(z)
is.integer(z)

class(4L * 2)
class(6L / 2L)

as.integer("12.35")
as.integer("Something")
```

## Numeric
It handels integers, decimals, both positive and negative and zero.
```{r}
k = 1 # By default it is numeric
k
class(k)
is.numeric(k)

x = 10.5
x
class(x)
is.numeric(x)
```

## Complex
```{r}
z = 5 + 6i
z
class(z)
is.complex(z)

as.complex(4)
```

## Character
```{r}
s = "statistics"
s
class(s)
is.character(s)

as.character(123)

# Number of characters
nchar(s)
# Substring
substr(s, start = 1, stop = 4)
```

## Dates
R has numerous different types of dates.
```{r}
# Stores just data
today = as.Date("2000-11-20")
today
class(today)

as.numeric(today)

# Stores data and time
today = as.POSIXct("2000-11-20 17:42")
today
class(today)

as.Date(today)
as.numeric(today)
```

# Structures
Statistics is study of data. So we will learn how to enter and manipulate data into R.

* Vecotr
* List
* <span style="color:lightgrey">Array</span>
* Matrix
* Data Frame

## Vector
Collection of elements of all the **same type**. It keeps track of the order that the data is entered in.\
**Indexing** of the vector starts from **1** !!!
```{r}
x = c(1, 2, 3); x
y = c(5, 6, 7, 8, 9, 10); y
z = c(TRUE, FALSE); z
s = c("S", "t", "a", "t", "i", "s", "t", "i", "c", "s"); s
```

Operations on te elements of a vector\
R is **vectorized laguage**, so operations are applied to each element of the vector automatically, without the need to loop through the vector.
```{r}
x = c(1, 2, 3); x
x + 2
5 * x
x^2
sqrt(x)
length(x)
```

Operations on vectors with the same length
```{r}
x = c(1, 2, 3)
y = c(5, 6, 7)
x + y
y - x
x * y
x / y

```

If the two vectors are with **different length**, to make the operation we repeat the elements of the shorter vector and then make the operation.
```{r}
x = c(1, 2, 3)
y = c(5, 6, 7, 8, 9, 10)
x + y 
y - x
x * y
x / y
```

Taking elements from a vactor
```{r}
y = c(5, 6, 7, 8, 9, 10)
y[3]

y[c(3, 4)]
y[c(4, 2, 3)]

y[-3]

y[8]

y[9] = 11
y[1:9]
```

Different types of generating vectors
```{r}
1:17
1:-15

seq(1, 17, by = 2)

rep(1:8, 2)
rep(1:8, each = 2)
```

Some more examples
```{r}
x = 5:7; x
y = 2:5; y
y <= 3
x[y <= 3]
sum(y <= 3)
sum(x[y <= 3])
which(y >= 3)
x[which(y >= 3)]

z = 5:21; z
z[y <= 3] # Differnt length vectors !!!

```

## Lists
Hold objects **not** necessarily from the **same type** or same **length**.
```{r}
x = list(1, 2, 3); x

y = list(a = 1, b = 2, c = 3); y

z = list(a = c(1, 2, 4, 5),
         b = c(TRUE, FALSE, TRUE),
         c = c("Statistics")); z
```

## Array
**Multidimentional** vector. All the elements must be of the **same type**.
```{r}
x = array(1:12, dim = c(2, 3, 2)); x
```

Taking elements of the array
```{r}
x[1, , ]

x[1, , 1]

x[, , 1]
```

## Matrix
All of the elements of the matrix must be of the **same type**.

Generating matrix
```{r}
A = matrix(c(2, 4, 1, 5, 7, 6),
           nrow = 2,
           ncol = 3,
           byrow = TRUE)
A
B = matrix(c(2, 4, 3, 1, 5, 7),
           nrow = 3,
           ncol = 2)
B
```

Dimentions of the matrix
```{r}
dim(B)
```

Taking elements of the matrix
```{r}
A[2, 3]
A[2, ]
A[, 3]
A[, c(1, 3)]
```

Operations on matrixes
```{r}
A + 3

G = matrix(c(rep(1:2, 3)), nrow = 2); G
A + G
A*G
t(A) # Transpose
G %*% t(A) # Multiplying two matrixes
```

You can also see:
```{r}
?cbind
?rbind
?dimnames
```

## Data Frame
Hold objects **not** necessarily from the **same type** but with the **same length**.
```{r}
first = 1:10 
second = c("aa", "bb", "cc", "dd", "ee", "ff", "gg", "hh", "ii", "jj") 
third = c(TRUE, FALSE, TRUE, TRUE, FALSE, TRUE, FALSE, FALSE, TRUE, FALSE) 

df = data.frame(first, second, third); df
View(df)
```

```{r}
class(df)
```


We can see the first or the last rows of the data frame
```{r}
head(df)
head(df, n = 3)
tail(df)
tail(df, n = 3)
```


We can see the dimensions of the matrix
```{r}
dim(df)
dim(df)[1]
dim(df)[2]

nrow(df)

ncol(df)
```

We can give or change the names of the columns of the data frame
```{r}
df = data.frame(A = first, B = second, C = third); df
View(df)
```

We can see the names of the columns
```{r}
names(df)
```

We can give names to the rows of the data frame
```{r}
rownames(df) = c("one", "two", "three", "four", "five", "six", "seven", "eight", "nine", "ten"); df
View(df)
```

Taking elements of the data frame
```{r}
 # row 2, column 3
df[2, 3]

# row 2
df[2, ]
# column 3
df[, 3]

# row 2, columns 2 and 3
df[2, 2:3]

# row 2, columns 1 and 2
df[2, c("A" , "B")]

# column 1
df$A

# column 2
df[[2]]
```

# Missing Data
* **NA** - missing value
* **NULL** - absence of anything
```{r}
x = c(1, 2, NA, 3, NA, 4); x
is.na(x)
sum(x)
sum(x, na.rm = TRUE) # Removing the NA values before the sum

y = c(1, NULL, 2); y
is.null(y)
length(y)
```

# Packages
## Installing packages
```{r}
?install.packages
help("install.packages")
# install.packages("UsingR")
# install.packages("MASS")
```

## Uninstlling packages
```{r}
# remove.packages("MASS")
```

## Loading packages
```{r}
?library
library(UsingR)
library(MASS)
```

## Unload packages
```{r}
# detach("package:MASS")
```

Taking information about a data frame, if it is available.
```{r}
mtcars
View(mtcars)
?mtcars
```

Some more examples with data frames from this packages
```{r}
# Dimentions of the data frame
dim(mtcars)
# Number of rows
nrow(mtcars)
# Number of columns
ncol(mtcars)

# row 1, column 2
mtcars[1, 2]
# row with name "Mazda RX4", column with name "cyl"
mtcars["Mazda RX4", "cyl"]
# coulmn 9
mtcars[[9]]
# column with name "am"
mtcars[["am"]]
# column with name "am"
mtcars$am
```

Attach data frame, so it is not needed to always write the data frame name infront of the column name.
```{r}
?attach

names(mtcars)
mtcars$mpg

attach(mtcars)
mpg
```

# Reading data from file

If the file is in your working directory, you do not need to specify the path to the file.

## Reading from .txt file
```{r}
?read.table
our.data = read.table("CData.txt", header = TRUE)
our.data
```

## Reading from .csv file
```{r}
?read.csv
pokemon = read.csv("pokemon.csv")
pokemon

titanic = read.csv("titanic.csv")
titanic
```

# Some more functions
* min() - min value
* max() - max value
* diff() - lagged differences
* cummin() - cumulative min
* cummax() - cumulative max
* sort() -
* order() -
```{r}
x = c(10, 1, 3, 4, 7, 34, 23, 10, 0, 0); x
min(x)
max(x)
diff(x)
cummin(x)
cummax(x)
```

# Statistics
## mean()
Arithmetic mean
\[
\bar{x} = \frac{1}{n} \sum_{i=1}^{n}x_{i} .
\]
```{r}
x = c(10, 1, 3, 4, 7, 34, 23, 10, 0, 0)
mean(x)

x = c(1, 1, 3, 4, 7, 3, 2, 1, 0, 0, 11000)
mean(x)
```
## median()
Sample median. 
```{r}
?median
```
## var()
\[
\sigma^2 = \frac{1}{n - 1} \sum_{i=1}^{n}(x_{i} - \bar{x})^2 .
\]
```{r}
?median
```
## sd()
\[
\sigma = \sqrt{\sigma^2} .
\]
```{r}
?median
```

# Examples
## Example 1:
We have the number of phone calls recorded for 2 weeks.
```{r}
phone.calls1 = c(10, 4, 1, 13, 7, 14, 8)
phone.calls1
phone.calls2 = c(8, 2, 4, 15, 3, 7, 15)
phone.calls2

phone.calls2[c(1, 2, 3, 4, 5)] # slicing
phone.calls2[1:5] # slicing
phone.calls2[seq(1,5,by = 1)] # slicing
```
But we have mistaken the number of calls in the first day of the second week.
```{r}
phone.calls2[1] = 9
phone.calls2
```
Print the number of phone calls in the fourth day of the second week.
```{r}
phone.calls2[4]
```
Print the number of phone calls in all the days, except the fourth day of the second week.
```{r}
phone.calls2[-4] # Gives everything except these indices
```
Print maximum number of phone calls for the second week
```{r}
max(phone.calls2)
```
Which days the phone calls were 15 in the second week
```{r}
phone.calls2 == 15
```
This gives a vector of TRUE and FALSE
How can we get the indices, so we can understand which day was that?
```{r}
which(phone.calls2 == 15) # extraction by a logical vector
```

Whe can make it one one line - but it is not readable
```{r}
(1:length(phone.calls2))[phone.calls2 == max(phone.calls2)]
```

Find the number of all phone calls for the second week.
```{r}
sum(phone.calls2)
```
Find the number of days with more than 5 calls in the second week
```{r}
sum(phone.calls2 > 5)
```
Find the differences in number of calls for day between the 2 weeks
```{r}
phone.calls1 - phone.calls2
```
Find the differences in the calls between the days during the second weeks
```{r}
diff(phone.calls2)
```

## Example 2:
We have the stock price for 10 days
```{r}
stock.price = c(45, 43, 46, 48, 51, 46, 50, 47, 46, 45)
stock.price
```
Find the min price and the max price of the stock during this 10 days
```{r}
max(stock.price)
min(stock.price)
```
Find the mean price of the stock during 10 days
```{r}
mean(stock.price)
```
Find the median of the stock price during this 10 days
```{r}
median(stock.price)
```
Lets add the price for some more days
```{r}
stock.price = c(stock.price, 48, 49, 51, 50, 49)
```
How many days we have been observing the stox price
```{r}
length(stock.price)
```
We have observed that on day 16 the price was 41. Can you add it?
```{r}
stock.price[16] = 41 # add to a specified index
```
Can you add the stock prices for the next 4 days?
```{r}
stock.price[17:20] = c(40, 38, 35, 40) # add to many specified indices
```
What was the max and min price at any moment for the stock
```{r}
cummax(stock.price)
cummin(stock.price)
```
What was the differences between everyday prices?
```{r}
diff(stock.price)
```

# Problems - Verzani - page 7
## Problem 2.1 
Suppose you keep track of your mileage each time you fill up. At your last 6 fill-ups the mileage was
\[
65311 \quad 65624 \quad 65908 \quad 66219 \quad 66499 \quad 66821 \quad 67145 \quad 67447
\]
Enter these numbers into R. Use the function diff on the data. What does it give?
```{r}
miles = c(65311, 65624, 65908, 66219, 66499, 66821, 67145, 67447)
x = diff(miles)
```
You should see the number of miles between fill-ups. Use the max to find the maximum number of miles between fill-ups, 
the mean function to find the average number of miles and the min to get the minimum number of miles.
```{r}
max(miles)
mean(miles)
min(miles)
```

## Problem 2.2
Suppose you track your commute times for two weeks (10 days) and you find the following times in minutes
\[
17 \quad 16 \quad 20 \quad 24 \quad 22 \quad 15 \quad 21 \quad 15 \quad 17 \quad 22
\]
Enter this into R.
```{r}
commutes = c(17, 16, 20, 24, 22, 15, 21, 15, 17, 22)
```
Use the function max to find the longest commute time, 
```{r}
max(commutes)
```
the function mean to find the average
```{r}
mean(commutes)
```
and the function min to find the minimum.
```{r}
min(commutes)
```
Oops, the 24 was a mistake. It should have been 18. How can you fix this?
```{r}
mistaken = which(commutes == 24)
commutes[mistaken] = 18
```
Do so, and then find the new average.
```{r}
mean(commutes)
```
How many times was your commute 20 minutes or more?
```{r}
sum(commutes >= 20)
```
What do you get? What percent of your commutes are less than 17 minutes? How can you answer this with R?
```{r}
sum(commutes < 17)/length(commutes)*100
```

## Problem 2.3
Your cell phone bill varies from month to month. Suppose your year has the following monthly amounts
\[
46 \quad 33 \quad 39 \quad 37 \quad 46 \quad 30 \quad 48 \quad 32 \quad 49 \quad 35 \quad 30 \quad 48
\]
Enter this data into a variable called bill.
```{r}
bill = c(46, 33, 39, 37, 46, 30, 48, 32, 49, 35, 30, 48)
```
Use the sum command to find the amount you spent this year on the cell phone.
```{r}
sum(bill)
```
What is the smallest amount you spent in a month?
```{r}
min(bill)
```
What is the largest?
```{r}
max(bill)
```
How many months was the amount greater than $40?
```{r}
sum(bill > 40)
```
What percentage was this?
```{r}
sum(bill > 40)/length(bill) * 100
```

## Problem 2.4
You want to buy a used car and find that over 3 months of watching the classifieds you see the following prices (suppose the cars are all similar)
\[
9000 \quad 9500 \quad 9400 \quad 9400 \quad 10000 \quad 9500 \quad 10300 \quad 10200
\]
```{r}
price = c(9000, 9500, 9400, 9400, 10000, 9500, 10300, 10200)
```
Use R to find the average value and compare it to Edmund's (http://www.edmunds.com) estimate of $9500.
```{r}
mean(price)
mean(price) > 9500
```
Use R to find the minimum value and the maximum value.
```{r}
min(price)
max(price)
```

## Problem 2.5
Try to guess the results of these R commands. We assume:
```{r}
x = c(1,3,5,7,9)
y = c(2,3,5,7,11,13)
```
* 1
```{r}
x + 1
```
* 2
```{r}
y*2
```
* 3
```{r}
length(x)
length(y)
```
* 4
```{r}
x + y
```
* 5
```{r}
sum(x > 5)
sum(x[x > 5])
```
* 6
```{r}
sum(x > 5 | x < 3)
```
* 7
```{r}
y[3]
```
* 8
```{r}
y[-3]
```
* 9
```{r}
y[x]
```
* 10
```{r}
y[y >= 7]
```

# Tasks - moodle

## Задача 1
Създайте вектор {8, 3, 8, 7, 15, 9, 12, 4, 9, 10}. Превърнете вектора в матрица 5x2. Задайте имена на редовете на матрицата - r1, r2 ... Добавте още една колона съдържаща нечетните числа - 1, 3, 5 ... Сортирайте матрицата по първа колона в растящ ред
```{r}
data = c(8, 3, 8, 7, 15, 9, 12, 4, 9, 10, 5, 1); data
matrix = matrix(data = data, nrow = 5, ncol = 2); matrix
matrix = matrix(data = data,
                nrow = 5,
                ncol = 2,
                dimnames = list(c("r1", "r2", "r3", "r4", "r5")))
matrix
matrix = cbind(matrix, c(1, 3, 5, 7, 9)); matrix
# Сортира първия стълб само
?sort # sort - Sort a vector or factor (partially) into ascending or descending order.
sort(matrix[,1])
# Сортира матрицата по първия стълб
?order # order - Returns a permutation which rearranges its first argument into ascending or descending order, breaking ties by further arguments.
matrix[order(matrix[,1]),]
```

## Задача 2
Разгледайте данните 'homedata' от пакета 'UsingR'. Създайте два вектора съдържащи цените на къщите през 1970 и 2000г.\
```{r}
# install.packages("UsingR")
library(UsingR)
homedata
head(homedata)
homedata1970 = homedata$y1970
head(homedata1970)
homedata2000 = homedata$y2000
head(homedata2000)
```
Определете:
а) Най-скъпата и най-евтината къща през 2000г, техните цени през 1970г
```{r}
minPrice = min(homedata2000); minPrice
maxPrice = max(homedata2000); maxPrice
homedata1970[which(homedata2000 == minPrice)]
homedata1970[which(homedata2000 == maxPrice)]
```
б) цените на 5-те най-скъпи къщи през 2000г
```{r}
tail(sort(homedata2000),5)
```
в) броят на къщите по-скъпи от 750 000 през 2000г
```{r}
sum(homedata2000 > 750000)
```
* изкарайте цените им
```{r}
mostExpensive = homedata2000[homedata2000 > 750000]
mostExpensive
```
г) средната цена през 1970г на къщите от в)
```{r}
mean(mostExpensive)
```
д) цената през 2000г на тези къщи, чиято цена е намаляла
```{r}
homedata2000[which(homedata2000 < homedata1970)]
```
е) 10-те къщи с най-голямо процентно увеличение на цената
```{r}
x = (homedata2000 - homedata1970)/homedata1970*100
sort(x, decreasing = TRUE)
head(sort(x, decreasing = TRUE), 10)
```
Имаме 1 къща с процентно увеличение на цената Inf, защото цената е през 1970та е 0
```{r}
sum(homedata1970 == 0)
```

## Задача 3
Разгледайте данните ’survey’ от пакета ’MASS’. Намерeте:
```{r}
# install.packages("MASS")
library(MASS)
?survey
# This data frame contains the responses of 237 Statistics I students at
# the University of Adelaide to a number of questions.
head(survey)
summary(survey)
```
а) броя на мъжете;
```{r}
?sum
sum(survey$Sex == "Male", na.rm = TRUE)
```
б) броя на мъжете пушачи;
```{r}
sum(survey$Sex == "Male" & survey$Smoke != "Never", na.rm = TRUE)
```
в) средната височина на мъжете;
```{r}
mean(survey$Height, na.rm = TRUE)
```
г) височината и пола на 6-те най-млади студента.
```{r}
head(survey[survey$Age == sort(survey$Age), c("Height", "Sex")], 6)
```

